Generating Random Logic Programs Using Constraint Programming

### 2020-08-13 seminar_talk/talk.tex
Example programs

% TODO: conditional independence?
% TODO: how is a predicate with its arguments represented?
% TODO: a full(er) example of independence propagation
% TODO: could talk about what algorithms are behind the negative cycle detection
% TODO: for the bigraph talk, I had 28 slides full info and it still wasn't
% enough. Maybe I should aim for at least 30. Right now I have 26.
% TODO: more constraints
% TODO: more motivation and related work
\begin{frame}{Variable Symmetry Breaking}
  \begin{block}{The Problem}
    Let $\{ \textcolor{variable}{\mathsf{W}}, \textcolor{variable}{\mathsf{X}},
    \textcolor{variable}{\mathsf{Y}} \}$ be the set of variables. Then
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{Y}},
      \textcolor{variable}{\mathsf{X}})
    \]
    is equivalent to
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{X}},
      \textcolor{variable}{\mathsf{Y}})
    \]
    and to
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{W}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{X}},
      \textcolor{variable}{\mathsf{W}})
    \]
  \end{block}
\end{frame}

@@

% TODO: all the extra plots and example I had in the appendix
\begin{frame}{Stratification and Negative Cycles}
  \[
    \textcolor{probability}{0.1} \prob
    \textcolor{predicate}{\mathtt{friend}}(\textcolor{variable}{\mathsf{X}},
    \textcolor{variable}{\mathsf{Y}}) \ifff
    \negg\textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}).
  \]
  \vfill
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \setlength{\leftmargini}{0pt}
      \begin{itemize}
      \item<3-> \alert{Stratum 1}: predicates with no negation
      \item<3-> \alert{Stratum 2}: new predicates s.t. negative literals refer to
        predicates in \alert{Stratum~1}
      \end{itemize}
      \vspace{5em}
    \end{column}
    \begin{column}{0.6\textwidth}
      \hspace{-6.92em}
      \begin{tikzpicture}
        \node[draw,ellipse] (friend) {\textcolor{predicate}{friend}};
        \node[draw,ellipse,right=of friend] (influences) {\textcolor{predicate}{influences}};
        \node[draw,ellipse,below=of influences] (stress) {\textcolor{predicate}{stress}};
        \node[draw,ellipse,below=of stress] (spont) {\textcolor{predicate}{cancer\_spont}};
        \node[draw,ellipse,below=of spont] (smoke) {\textcolor{predicate}{cancer\_smoke}};
        \node[draw,ellipse,left=of spont] (person) {\textcolor{predicate}{person}};
        \node[draw,ellipse,right=of spont] (cancer) {\textcolor{predicate}{cancer}};
        \node[draw,ellipse,above=of cancer] (smokes) {\textcolor{predicate}{smokes}};
        \draw[-{Latex[length=2mm]}] (friend) -- (influences);
        \draw[-{Latex[length=2mm]}] (influences) -- (smokes);
        \draw[-{Latex[length=2mm]}] (smokes) -- (cancer);
        \draw[-{Latex[length=2mm]}] (stress) -- (smokes);
        \draw[-{Latex[length=2mm]}] (spont) -- (cancer);
        \draw[-{Latex[length=2mm]}] (smoke) -- (cancer);
        \draw[-{Latex[length=2mm]}] (person) -- (stress);
        \draw[-{Latex[length=2mm]}] (person) -- (spont);
        \draw[-{Latex[length=2mm]}] (person) -- (smoke);
        \draw<2->[-{Latex[length=2mm]},draw=red] (smokes) -- (friend);
      \end{tikzpicture}
    \end{column}
  \end{columns}
\end{frame}

@@

>>>>

% TODO: say something about heuristics
% TODO: conditional independence?
% TODO: how is a predicate with its arguments represented?
% TODO: a full(er) example of independence propagation
% TODO: could talk about what algorithms are behind the negative cycle detection
% TODO: for the bigraph talk, I had 28 slides full info and it still wasn't
% enough. Maybe I should aim for at least 30. Right now I have 26.
% TODO: more constraints
% TODO: more motivation and related work
\begin{frame}{Variable Symmetry Breaking}
  \begin{block}{The Problem}
    Let $\{ \textcolor{variable}{\mathsf{W}}, \textcolor{variable}{\mathsf{X}},
    \textcolor{variable}{\mathsf{Y}} \}$ be the set of variables. Then
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{Y}},
      \textcolor{variable}{\mathsf{X}})
    \]
    is equivalent to
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{X}},
      \textcolor{variable}{\mathsf{Y}})
    \]
    and to
    \[
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{W}})
      \ifff
      \textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{X}}),
      \textcolor{predicate}{\mathtt{influences}}(\textcolor{variable}{\mathsf{X}},
      \textcolor{variable}{\mathsf{W}})
    \]
  \end{block}
\end{frame}

@@

\begin{frame}{Stratification and Negative Cycles}
  \[
    \textcolor{probability}{0.1} \prob
    \textcolor{predicate}{\mathtt{friend}}(\textcolor{variable}{\mathsf{X}},
    \textcolor{variable}{\mathsf{Y}}) \ifff
    \negg\textcolor{predicate}{\mathtt{smokes}}(\textcolor{variable}{\mathsf{Y}}).
  \]
  \vfill
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \setlength{\leftmargini}{0pt}
      \begin{itemize}
      \item<3-> \alert{Stratum 1}: predicates with no negation
      \item<3-> \alert{Stratum 2}: new predicates s.t. negative literals refer to
        predicates in \alert{Stratum~1}
      \end{itemize}
      \vspace{5em}
    \end{column}
    \begin{column}{0.6\textwidth}
      \hspace{-6.92em}
      \begin{tikzpicture}
        \node[draw,ellipse] (friend) {\textcolor{predicate}{friend}};
        \node[draw,ellipse,right=of friend] (influences) {\textcolor{predicate}{influences}};
        \node[draw,ellipse,below=of influences] (stress) {\textcolor{predicate}{stress}};
        \node[draw,ellipse,below=of stress] (spont) {\textcolor{predicate}{cancer\_spont}};
        \node[draw,ellipse,below=of spont] (smoke) {\textcolor{predicate}{cancer\_smoke}};
        \node[draw,ellipse,left=of spont] (person) {\textcolor{predicate}{person}};
        \node[draw,ellipse,right=of spont] (cancer) {\textcolor{predicate}{cancer}};
        \node[draw,ellipse,above=of cancer] (smokes) {\textcolor{predicate}{smokes}};
        \draw[-{Latex[length=2mm]}] (friend) -- (influences);
        \draw[-{Latex[length=2mm]}] (influences) -- (smokes);
        \draw[-{Latex[length=2mm]}] (smokes) -- (cancer);
        \draw[-{Latex[length=2mm]}] (stress) -- (smokes);
        \draw[-{Latex[length=2mm]}] (spont) -- (cancer);
        \draw[-{Latex[length=2mm]}] (smoke) -- (cancer);
        \draw[-{Latex[length=2mm]}] (person) -- (stress);
        \draw[-{Latex[length=2mm]}] (person) -- (spont);
        \draw[-{Latex[length=2mm]}] (person) -- (smoke);
        \draw<2->[-{Latex[length=2mm]},draw=red] (smokes) -- (friend);
      \end{tikzpicture}
    \end{column}
  \end{columns}
\end{frame}

@@

\section{Example Programs}

\begin{frame}{One-Liners}
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{block}{Setup}
        \setlength{\leftmargini}{7pt}
        \begin{itemize}
        \item predicate $\textcolor{predicate}{\mathtt{p}}/1$
        \item variable $\textcolor{variable}{\mathsf{X}}$
        \item no constants
        \item \alert{1} clause
        \item \alert{4} nodes
        \item no negative cycles
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.6\textwidth}
      \begin{block}{(All) Programs}
        \begin{itemize}
        \item
          $\textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}}).$
        \item $\textcolor{probability}{0.7} \prob
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})
          \ifff
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}}).$
        \item $\textcolor{probability}{0.8} \prob
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})
          \ifff
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}});
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}}).$
        \item $\textcolor{probability}{0.7} \prob
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})
          \ifff
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}}),
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}}).$
        \item $\textcolor{probability}{0.1} \prob
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})
          \ifff
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}});
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}});
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}}).$
        \item $\textcolor{probability}{0.8} \prob
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}})
          \ifff
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}}),
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}}),
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}}).$
        \end{itemize}
      \end{block}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Symmetry Breaking in Action}
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{block}{Setup}
        \setlength{\leftmargini}{7pt}
        \begin{itemize}
        \item predicate $\textcolor{predicate}{\mathtt{p}}/3$
        \item variables: $\textcolor{variable}{\mathsf{X}}$,
          $\textcolor{variable}{\mathsf{Y}}$, $\textcolor{variable}{\mathsf{Z}}$
        \item no constants
        \item \alert{1} clause
        \item \alert{1} node
        \item no cycles at all
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.6\textwidth}
      \begin{block}{(All) Programs}
        \begin{itemize}
        \item $\textcolor{probability}{0.8} \prob
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{Z}},
          \textcolor{variable}{\mathsf{Z}}, \textcolor{variable}{\mathsf{Z}}).$
        \item
          $\textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{Y}},
          \textcolor{variable}{\mathsf{Y}}, \textcolor{variable}{\mathsf{Z}}).$
        \item
          $\textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{Y}},
          \textcolor{variable}{\mathsf{Z}}, \textcolor{variable}{\mathsf{Y}}).$
        \item $\textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{Y}},
          \textcolor{variable}{\mathsf{Z}}, \textcolor{variable}{\mathsf{Z}}).$
        \item $\textcolor{probability}{0.1} \prob
          \textcolor{predicate}{\mathtt{p}}(\textcolor{variable}{\mathsf{X}},
          \textcolor{variable}{\mathsf{Y}}, \textcolor{variable}{\mathsf{Z}}).$
        \end{itemize}
      \end{block}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{A Larger Example}
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{block}{Setup}
        \setlength{\leftmargini}{7pt}
        \begin{itemize}
        \item predicates: $\textcolor{predicate}{\mathtt{p}}/1$,
          $\textcolor{predicate}{\mathtt{q}}/2$,
          $\textcolor{predicate}{\mathtt{r}}/3$
        \item variables: $\textcolor{variable}{\mathsf{X}}$,
          $\textcolor{variable}{\mathsf{Y}}$, $\textcolor{variable}{\mathsf{Z}}$
        \item constants: $\textcolor{constant}{\mathit{a}}$,
          $\textcolor{constant}{\mathit{b}}$, $\textcolor{constant}{\mathit{c}}$
        \item \alert{5} clauses
        \item \alert{5} nodes
        \item no negative cycles
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.6\textwidth}
      \begin{block}{A Random Program}
        \begin{align*}
          &\textcolor{predicate}{\mathtt{p}}(\textcolor{constant}{\mathit{b}}) \ifff \negg(\textcolor{predicate}{\mathtt{q}}(\textcolor{constant}{\mathit{a}}, \textcolor{constant}{\mathit{b}}), \textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{X}}, \textcolor{variable}{\mathsf{Y}}), \textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{Z}}, \textcolor{variable}{\mathsf{X}})). \\
          \textcolor{probability}{0.4} \prob &\textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{X}}, \textcolor{variable}{\mathsf{X}}) \ifff \negg \textcolor{predicate}{\mathtt{r}}(\textcolor{variable}{\mathsf{Y}}, \textcolor{variable}{\mathsf{Z}}, \textcolor{constant}{\mathit{a}}). \\
          &\textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{X}}, \textcolor{constant}{\mathit{a}}) \ifff \textcolor{predicate}{\mathtt{r}}(\textcolor{variable}{\mathsf{Y}}, \textcolor{variable}{\mathsf{Y}}, \textcolor{variable}{\mathsf{Z}}). \\
          &\textcolor{predicate}{\mathtt{q}}(\textcolor{variable}{\mathsf{X}}, \textcolor{constant}{\mathit{a}}) \ifff \textcolor{predicate}{\mathtt{r}}(\textcolor{variable}{\mathsf{Y}}, \textcolor{constant}{\mathit{b}}, \textcolor{variable}{\mathsf{Z}}). \\
          &\textcolor{predicate}{\mathtt{r}}(\textcolor{variable}{\mathsf{Y}}, \textcolor{constant}{\mathit{b}}, \textcolor{variable}{\mathsf{Z}}).
        \end{align*}
      \end{block}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Examples of Predicate Independence}
  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{block}{Setup}
        \setlength{\leftmargini}{7pt}
        \begin{itemize}
        \item predicates: $\textcolor{predicate}{\mathtt{p}}/1$,
          $\textcolor{predicate}{\mathtt{q}}/1$,
          $\textcolor{predicate}{\mathtt{r}}/1$
        \item no variables
        \item constant $\textcolor{constant}{\mathit{a}}$
        \item \alert{3} clauses
        \item \alert{3} nodes
        \item no negative cycles
        \item $\textcolor{predicate}{\mathtt{p}} \independent
          \textcolor{predicate}{\mathtt{q}}$
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.6\textwidth}
      \begin{block}{A Few Random Programs}
        \begin{itemize}
          \setlength\itemsep{1em}
        \item $\begin{aligned}[t]
            \textcolor{probability}{0.5} \prob
            &\textcolor{predicate}{\mathtt{p}}(\textcolor{constant}{\mathit{a}})
            \ifff
            \textcolor{predicate}{\mathtt{p}}(\textcolor{constant}{\mathit{a}});
            \textcolor{predicate}{\mathtt{p}}(\textcolor{constant}{\mathit{a}}).
            \\ \textcolor{probability}{0.2} \prob
            &\textcolor{predicate}{\mathtt{q}}(\textcolor{constant}{\mathit{a}})
            \ifff
            \textcolor{predicate}{\mathtt{q}}(\textcolor{constant}{\mathit{a}}),
            \textcolor{predicate}{\mathtt{q}}(\textcolor{constant}{\mathit{a}}).
            \\ \textcolor{probability}{0.4} \prob
            &\textcolor{predicate}{\mathtt{r}}(\textcolor{constant}{\mathit{a}})
            \ifff \negg
            \textcolor{predicate}{\mathtt{q}}(\textcolor{constant}{\mathit{a}}).
          \end{aligned}$
        \item $\begin{aligned}[t]
            &\textcolor{predicate}{\mathtt{p}}(\textcolor{constant}{\mathit{a}})
            \ifff
            \textcolor{predicate}{\mathtt{p}}(\textcolor{constant}{\mathit{a}}).
            \\ \textcolor{probability}{0.5} \prob
            &\textcolor{predicate}{\mathtt{q}}(\textcolor{constant}{\mathit{a}})
            \ifff
            \textcolor{predicate}{\mathtt{r}}(\textcolor{constant}{\mathit{a}});
            \textcolor{predicate}{\mathtt{q}}(\textcolor{constant}{\mathit{a}}).
            \\
            &\textcolor{predicate}{\mathtt{r}}(\textcolor{constant}{\mathit{a}})
            \ifff
            \textcolor{predicate}{\mathtt{r}}(\textcolor{constant}{\mathit{a}});
            \textcolor{predicate}{\mathtt{r}}(\textcolor{constant}{\mathit{a}}).
          \end{aligned}$
        \item $\begin{aligned}[t]
            &\textcolor{predicate}{\mathtt{p}}(\textcolor{constant}{\mathit{a}})
            \ifff
            \textcolor{predicate}{\mathtt{p}}(\textcolor{constant}{\mathit{a}});
            \textcolor{predicate}{\mathtt{p}}(\textcolor{constant}{\mathit{a}}).
            \\ \textcolor{probability}{0.6} \prob
            &\textcolor{predicate}{\mathtt{q}}(\textcolor{constant}{\mathit{a}})
            \ifff
            \textcolor{predicate}{\mathtt{q}}(\textcolor{constant}{\mathit{a}}).
            \\ \textcolor{probability}{0.7} \prob
            &\textcolor{predicate}{\mathtt{r}}(\textcolor{constant}{\mathit{a}})
            \ifff \negg
            \textcolor{predicate}{\mathtt{q}}(\textcolor{constant}{\mathit{a}}).
          \end{aligned}$
        \end{itemize}
      \end{block}
    \end{column}
  \end{columns}
\end{frame}
